---
title: "Storm Data"
author: "Diana Marcela Rosero"
date: "26/7/2020"
output: html_document
---

## Impact of Severe Weather Events on Public Health and Economy in the United States

### Synonpsis  
In this report, we aim to analyze the impact of different weather events on public health and economy based on the storm database collected from the U.S. National Oceanic and Atmospheric Administration's (NOAA) from 1950 - 2011. We will use the estimates of fatalities, injuries, property and crop damage to decide which types of event are most harmful to the population health and economy. From these data, we found that excessive heat and tornado are most harmful with respect to population health, while flood, drought, and hurricane/typhoon have the greatest economic consequences.

### Basic settings
```{r, echo = TRUE}
# Import data
setwd("E:/ESCRITORIO/Desktop/specdata")
if(!exists("storm.data")) {
    storm.data <- read.csv(bzfile("repdata_data_StormData.csv.bz2"),header = TRUE)
  }
# Libraries
library(dplyr)
library(ggplot2)
library(plyr)
Sys.setlocale("LC_TIME", "English")

# some information about the variables
str(strom.data)
summary(strom.data)
dim(storm.data)
```

There are 902297 rows and 37 columns in total.
The events in the database start in the year 1950 and end in November 2011. In the earlier years of the database there are generally fewer events recorded, most likely due to a lack of good records. More recent years should be considered more complete.

# Extracting variables of interest for analysis of weather impact on health and economy
```{r, echo = TRUE}
vars <- c( "EVTYPE", "FATALITIES", "INJURIES", "PROPDMG", "PROPDMGEXP", "CROPDMG", "CROPDMGEXP")
mydata <- storm.data[, vars]
tail(mydata)
```

# Checking for missing values
```{r, echo=TRUE}
sum(is.na(mydata$FATALITIES))
sum(is.na(mydata$INJURIES))
# Check for missing values in economic variables for “size” of damage - there is no NA’s in the data.
sum(is.na(mydata$PROPDMG))
sum(is.na(mydata$CROPDMG))
# Check for missing values in economic variables for units damage - there is no NA’s in the data.
sum(is.na(mydata$PROPDMGEXP))
sum(is.na(mydata$CROPDMGEXP))
```

# Transforming extracted variables
```{r, echo=TRUE}
sort(table(mydata$EVTYPE), decreasing = TRUE)[1:10]
# create a new variable EVENT to transform variable EVTYPE in groups
mydata$EVENT <- "OTHER"
# group by keyword in EVTYPE
mydata$EVENT[grep("HAIL", mydata$EVTYPE, ignore.case = TRUE)] <- "HAIL"
mydata$EVENT[grep("HEAT", mydata$EVTYPE, ignore.case = TRUE)] <- "HEAT"
mydata$EVENT[grep("FLOOD", mydata$EVTYPE, ignore.case = TRUE)] <- "FLOOD"
mydata$EVENT[grep("WIND", mydata$EVTYPE, ignore.case = TRUE)] <- "WIND"
mydata$EVENT[grep("STORM", mydata$EVTYPE, ignore.case = TRUE)] <- "STORM"
mydata$EVENT[grep("SNOW", mydata$EVTYPE, ignore.case = TRUE)] <- "SNOW"
mydata$EVENT[grep("TORNADO", mydata$EVTYPE, ignore.case = TRUE)] <- "TORNADO"
mydata$EVENT[grep("WINTER", mydata$EVTYPE, ignore.case = TRUE)] <- "WINTER"
mydata$EVENT[grep("RAIN", mydata$EVTYPE, ignore.case = TRUE)] <- "RAIN"
# listing the transformed event types 
sort(table(mydata$EVENT), decreasing = TRUE)
# Checking the values for variables that represent units od dollars
sort(table(mydata$PROPDMGEXP), decreasing = TRUE)[1:10]
sort(table(mydata$CROPDMGEXP), decreasing = TRUE)[1:10]
```

There is some mess in units, so we transform those variables in one unit (dollar) variable by the following rule:
* K or k: thousand dollars (10^3)
* M or m: million dollars (10^6)
* B or b: billion dollars (10^9)
* the rest would be consider as dollars

New variable(s) is product of value of damage and dollar unit
```{r, echo=TRUE}
mydata$PROPDMGEXP <- as.character(mydata$PROPDMGEXP)
mydata$PROPDMGEXP[is.na(mydata$PROPDMGEXP)] <- 0 # NA's considered as dollars
mydata$PROPDMGEXP[!grepl("K|M|B", mydata$PROPDMGEXP, ignore.case = TRUE)] <- 0 # everything exept K,M,B is dollar
mydata$PROPDMGEXP[grep("K", mydata$PROPDMGEXP, ignore.case = TRUE)] <- "3"
mydata$PROPDMGEXP[grep("M", mydata$PROPDMGEXP, ignore.case = TRUE)] <- "6"
mydata$PROPDMGEXP[grep("B", mydata$PROPDMGEXP, ignore.case = TRUE)] <- "9"
mydata$PROPDMGEXP <- as.numeric(as.character(mydata$PROPDMGEXP))
mydata$property.damage <- mydata$PROPDMG * 10^mydata$PROPDMGEXP

mydata$CROPDMGEXP <- as.character(mydata$CROPDMGEXP)
mydata$CROPDMGEXP[is.na(mydata$CROPDMGEXP)] <- 0 # NA's considered as dollars
mydata$CROPDMGEXP[!grepl("K|M|B", mydata$CROPDMGEXP, ignore.case = TRUE)] <- 0 # everything exept K,M,B is dollar
mydata$CROPDMGEXP[grep("K", mydata$CROPDMGEXP, ignore.case = TRUE)] <- "3"
mydata$CROPDMGEXP[grep("M", mydata$CROPDMGEXP, ignore.case = TRUE)] <- "6"
mydata$CROPDMGEXP[grep("B", mydata$CROPDMGEXP, ignore.case = TRUE)] <- "9"
mydata$CROPDMGEXP <- as.numeric(as.character(mydata$CROPDMGEXP))
mydata$crop.damage <- mydata$CROPDMG * 10^mydata$CROPDMGEXP
```

```{r, echo=TRUE}
# Print of first 10 values for property damage (in dollars) that most appear in the data
sort(table(mydata$property.damage), decreasing = TRUE)[1:10]
# Print of first 10 values for crop damage (in dollars) that most appear in the data
sort(table(mydata$crop.damage), decreasing = TRUE)[1:10]
```

## Analysis
# Aggregating events for public health variables
Table of public health problems by event type
```{r, echo=TRUE}
# aggregate FATALITIES and INJURIES by type of EVENT
agg.fatalites.and.injuries <- ddply(mydata, .(EVENT), summarize, Total = sum(FATALITIES + INJURIES,  na.rm = TRUE))
agg.fatalites.and.injuries$type <- "fatalities and injuries"
  
# aggregate FATALITIES by type of EVENT
agg.fatalities <- ddply(mydata, .(EVENT), summarize, Total = sum(FATALITIES, na.rm = TRUE))
agg.fatalities$type <- "fatalities"

# aggregate INJURIES by type of EVENT
agg.injuries <- ddply(mydata, .(EVENT), summarize, Total = sum(INJURIES, na.rm = TRUE))
agg.injuries$type <- "injuries"

# combine all
agg.health <- rbind(agg.fatalities, agg.injuries)

health.by.event <- join (agg.fatalities, agg.injuries, by="EVENT", type="inner")
health.by.event
```

# Aggregating events for economic variables
```{r, echo=TRUE}
# aggregate PropDamage and CropDamage by type of EVENT
agg.propdmg.and.cropdmg <- ddply(mydata, .(EVENT), summarize, Total = sum(property.damage + crop.damage,  na.rm = TRUE))
agg.propdmg.and.cropdmg$type <- "property and crop damage"

# aggregate PropDamage by type of EVENT
agg.prop <- ddply(mydata, .(EVENT), summarize, Total = sum(property.damage, na.rm = TRUE))
agg.prop$type <- "property"

# aggregate INJURIES by type of EVENT
agg.crop <- ddply(mydata, .(EVENT), summarize, Total = sum(crop.damage, na.rm = TRUE))
agg.crop$type <- "crop"

# combine all
agg.economic <- rbind(agg.prop, agg.crop)


economic.by.event <- join (agg.prop, agg.crop, by="EVENT", type="inner")
economic.by.event
```

## Results
# Across the United States, which types of events are most harmful with respect to population health?
```{r, echo=TRUE}
# transform EVENT to factor variable for health variables
agg.health$EVENT <- as.factor(agg.health$EVENT)

# plot FATALITIES and INJURIES by EVENT
health.plot <- ggplot(agg.health, aes(x = EVENT, y = Total, fill = type)) + geom_bar(stat = "identity") +
  coord_flip() +
  xlab("Event Type") + 
  ylab("Total number of health impact") +
  ggtitle("Weather event types impact on public health") +
  theme(plot.title = element_text(hjust = 0.5))
print(health.plot)
```
The most harmful weather event for health (in number of total fatalites and injuries) is, by far, a tornado.

# Across the United States, which types of events have the greatest economic consequences?
```{r, echo=TRUE}
#transform EVENT to factor variable for economic variables
agg.economic$EVENT <- as.factor(agg.economic$EVENT)

# plot PROPERTY damage and CROP damage by EVENT
economic.plot <- ggplot(agg.economic, aes(x = EVENT, y = Total, fill = type)) + geom_bar(stat = "identity") +
  coord_flip() +
  xlab("Event Type") + 
  ylab("Total damage in dollars") +
  ggtitle("Weather event types impact on property and crop damage") +
  theme(plot.title = element_text(hjust = 0.5))
print(economic.plot) 
```
The most devastating weather event with the greatest economic cosequences (to property and crops) is a flood.

## Conclusion 
From these data, we found that **excessive heat** and **tornado** are most harmful with respect to population health, while **flood**, **drought**, and **hurricane/typhoon** have the greatest economic consequences.